---
title: "Climate Change Module"
author: 'Helen Choi'
format:
  html:
    embed-resources: true
---

```{r message=FALSE}
library(tidyverse)
```

## Warm-up: Examining CO2 trends in R

- Example from <http://climate.nasa.gov/vital-signs/carbon-dioxide/>
- Raw data from <https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt>

In 1958, Dr. Charles David Keeling (1928-2005), a scientist at Scripps Institute of Oceanography, began collecting data on atmospheric CO2 concentration at the Mauna Loa Observatory located in Hawaii. This dataset allowed us to understand the degree to which climate change is human-caused through our burning of fossil fuels and release of CO2 into the atmosphere. Due to his scientific achievements, Dr. Keeling was awarded the National Medal of Science by President George W. Bush in 2002. This is the highest award for lifetime scientific achievement that can be granted in the U.S. Today, you get to analyze this same dataset, except that you have more data than was available to Dr. Keeling and his colleagues because your dataset extends up to the current time.

To read the code, you will use a new function called `read_table`.  It is similar to `read_csv` except it looks for spaces between column entries rather than commas (remember that csv stands for comma-separated values).  Others include `read_tsv`, which uses tabs to separate entries.  You can discover the separation type by putting <https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt> into your web browser and examining the file.  The code also includes an argument for comment, which denotes the character used to define a line in the file as being a comment rather than data or a header.  The file also reveals that the column headers are on lines with a comment character, so they won't be read.  You will use the argument `col_names` to define the headers manually.  Finally, `NA` values are defined by the numbers -1 and -99, so they need to be defined using the `na` argument (otherwise, they will be read as numeric values).

```{r message=FALSE}
co2 <-  read_table("https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt", 
                  comment="#",
                  col_names = c("year", "month", "decimal_date",
                                "monthly_average",
                                "deseasonalized", "days", "sd_days",
                                "unc_month_mean"),
                  na = c("-1", "-99.99"))
co2
```



**Question 1:**

Describe the overall trend in the CO2 data.

**Answer 1:**
```{r}
ggplot(co2, aes(x = decimal_date, y = monthly_average)) + 
  geom_line() + 
  geom_line(aes(y = deseasonalized), color = "blue") +
  labs(x = "Year", y = "CO2 concentration (ppm)",
       title = "Overall Trend in CO2 Concentration over Time"
)
```

The overall trend in the CO2 data shows an increase in CO2 concentration over the time of 1960 to the present. This graph shows that CO2 concentration has been consistently rising over the years, this is likely due to more pollution from things such as human activity and the burning of fossil fuels.

**Question 2:**

How does CO2 vary within a year?  What month is it at max?  Min?  What might explain this sub-annual pattern? (you will need to write code and make a plot to answer this question)

**Answer 2:**

```{r}
monthly_co2 <- co2 |> 
  group_by(month) |> 
  summarize(mean_co2 = mean(monthly_average, na.rm = TRUE))

ggplot(monthly_co2, aes(x = month, y = mean_co2)) + 
  geom_line() +
  geom_point() +
  labs(x = "Month", y = "Average CO2 Concentration (ppm)",
       title = "Average Monthly CO2 Concentration") 
```

CO2 increases during the colder months(Fall and Winter) and decreases during the warmer months(Spring and Summer). The month CO2 concentration is at its maximum is in May and the month it is at its minimum is September. This decrease in concentration in warmer months might be explained by there being more plants absorbing CO2 through photosynthesis which decreases the amount in the atmosphere. In the colder months, the increase in CO2 might be explained by plants dying and losing their leaves, which causes less CO2 to be absorbed by these plants. Also, the decomposition of these dying planst can also cause more CO2 to be released in the atmosphere.


## Global Temperature Data

Current climate change affects many aspects of the environment, with socio-economic consequences. For example, a warmer climate can allow new diseases to be introduced and persist (e.g., West Nile became established in the United States after an unusually warm winter, which allowed the mosquitoes that carry the virus to survive and spread). We are concerned not only with the actual temperature but also with the rate at which it changes. Rapid changes make it more likely that species cannot adapt and will go extinct. 

Each of the most recent years has been the warmest on record. In this section, we will analyze global mean temperature data.

Data from: <https://climate.nasa.gov/vital-signs/global-temperature/>

**Question 3:**

Describe the data set to the best of your ability, given the documentation provided. 

- Where is the data from?
- Describe the class of each column and what units it is measured in. 
- What is the difference between "global temperature" and "global temperature anomaly"? 


**Answer 3:**
This dataset represents Global Land-Ocean Temperature Index on Earth. It is from NASA/GISS and shows information on global warming trends. The first column is year which is an integer as the years are represented in whole numbers. The second column is No_Smoothing which is the class numeric as the values are represented as decimal numbers. The units are in Celsius. The third column is Lowess(5) which is also the class numeric and the units are also in Celsius. The difference between global temperature and global temperature anomaly is that global temperature is the actual measured surface temperature at a given time. Global Temperature Anomaly is the difference between the measured surface temperature at a given time from a historical baseline temperature. 

**Question 4:**

Construct the necessary R code to import and prepare for plotting the following data set: <https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt>

You'll need to determine the file's delimiter and any comments or skips. You will also need to be sure that you have column names. You will not directly use the code above to read the CO2 data, but that code provides helpful tips for reading the data.

**Answer 4:**
```{r}
tempData <-  read_table("https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt", 
                  skip = 5,
                  col_names = c("Year", "No_Smoothing", "Lowess"))
head(tempData)
```

**Question 5:**

Plot the trend in global mean temperatures over time.  Describe what you see in the plot and how you interpret the patterns you observe.

**Answer 5:**
```{r}
# Raw Temperature Anomalies
ggplot(tempData, aes(x = Year, y = No_Smoothing)) + 
  geom_line(color = "blue") + 
  geom_point() +
  labs(x = "Year", y = "Temperature Anomaly",
       title = "Global Mean Temperature Anomalies Raw Data") 
# Smoothed Temperature Anomalies
ggplot(tempData, aes(x = Year, y = Lowess)) + 
  geom_line() + 
  geom_point() +
  labs(x = "Year", y = "Temperature Anomaly",
       title = "Global Mean Temperature Anomalies Smoothed Data") 
```

In the plot, I see a pattern of steady warming wth a significant increase in the amount of warming after around 1965. There are fluctuations in the data, but the long-term trend is showing an increase over time. This shows that global warming has been increasing significantly over the years which can be explained by things such as burning of fossil fuels/greenhouse has emissions.

## Evaluating the evidence for a "Pause" in warming?

The [2013 IPCC Report](https://www.ipcc.ch/pdf/assessment-report/ar5/wg1/WG1AR5_SummaryVolume_FINAL.pdf) included a tentative observation of a "much smaller increasing trend" in global mean temperatures since 1998 than was observed previously.  This led to much discussion in the media about the existence of a "Pause" or "Hiatus" in global warming rates, as well as much research looking into where the extra heat could have gone.  (Examples discussing this question include articles in [The Guardian](http://www.theguardian.com/environment/2015/jun/04/global-warming-hasnt-paused-study-finds), [BBC News](http://www.bbc.com/news/science-environment-28870988), and [Wikipedia](https://en.wikipedia.org/wiki/Global_warming_hiatus)). 

You will use rolling averages to help you explore the evidence for a pause. Since you have not been provided instructions for calculating rolling means, the learning objective of this question is to practice finding the solution.

**Question 6:**

Use a search engine (e.g., Google) or a chat LLM to find out how to calculate a rolling average in R. What search term or chat LLM prompt did you use?  What website or answer from the chat LLM did you end up using?  How much did you need to modify the code from the website or chat LLM to answer the question successfully?

**Answer 6:**
```{r}
library(zoo)
rollingAvg <- rollmean(tempData$No_Smoothing, k = 7, align = "right", fill = NA)
rollingAvg
```

To answer this question I just looked up in google "How to calculate a rolling average in R". I found a good explanation on the GeeksforGeeks website: https://www.geeksforgeeks.org/how-to-calculate-a-rolling-average-in-r/
that provided a good example using the rollmean() function from the zoo package:
rolling_avg <- rollmean(temperature_xts, k = 7, align = "right", fill = NA).
I did not need to modify my code too much, only changing the dataset that is being used.

**Question 7:**

- What is the meaning of "5-year average" vs. "annual average"?
- Create a data frame from the annual temperature anomaly data (from Question 4) with three new columns: 5-year running averages, 10-year running averages, and 20-year running averages.

**Answer 7:**
An annual average represents the temperature anomaly for a single year. It shows how much that year's temperature deviated from a baseline temperature. A 5-year average is a rolling mean that averages temperature anomalies of the previous 5 years and how that changed from a baseline temperature.

```{r}
tempData2 <- tempData |> 
  select(Year, No_Smoothing) |> 
  mutate(
    running5 = rollmean(No_Smoothing, k = 5, align = "right", fill = NA),
    running10 = rollmean(No_Smoothing, k = 10, align = "right", fill = NA),
    running20 = rollmean(No_Smoothing, k = 20, align = "right", fill = NA)
  )
head(tempData2)
```

**Question 8:**

Plot the different averages on the *same plot* and describe what differences you see and why.  

**Answer 8:**
```{r}
ggplot(tempData2, aes(x = Year)) +
  geom_line(aes(y = running5), color = "blue") +
  geom_line(aes(y = running10), color = "red") +
  geom_line(aes(y = running20), color = "green") +
  labs(x = "Year", y = "Temperature Anomaly",
       title = "Rolling Averages of Global Mean Temperature Anomalies")
```

Legend: 5-year running averages: Blue, 10-year running averages: Red, 20-year running averages: Green

The differences I see between these plots, is the larger the time window, the smoother the line becomes. The green line(20 years) is the smoothest and the blue line(5 years) shows more variability. This is because the longer running averages cause less fluctuations which makes long term trends easier to see.

**Question 9:**

By examining the data here, what evidence do you find or not find for such a pause? 

**Answer 9:**
By examining the data, there is little evidence to indicate that there was a significant pause in global warming based on the three rolling averages. The long-term indicates that the trend of temperature is consistently increasing indicated by the upward moving line. The pause that could have been seen could have just been explained by things such as variability in climate at Earth.

## Longer term trends in CO2 Records

When analyzing Earth’s climate, it is important to remember that Earth is 4.54 billion years old. Our analyses so far have only looked at recent history. How can we compare the recent data to prehistoric times? Are the current rates of change similar or different from those the earth has experienced in the past? To explore this, we can use data from ice cores drilled at the poles. 

Hundreds of ice cores have been extracted from polar ice because they contain valuable data on atmospheric chemistry over pre-historic times. These valuable data exist in tiny air bubbles trapped in the ice. These air bubbles contain the same gases in the same ratios as the atmosphere at the time when the ice formed. The data you will analyze today are from ice cores extracted from the Vostok research station in Antarctica. As you have probably assumed, the depth of the ice core is related to how old the ice is; deep ice is older. There are two other variables that you will be analyzing from the ice cores.  You will analyze CO2 concentration, which has been measured from air bubbles trapped in the ice. We can use these data to see what rates of change were like during this pre-historic period, during which human activity was minimal. 

[Ice core data](https://data.ess-dive.lbl.gov/view/doi%3A10.3334%2FCDIAC%2FATG.009):

Vostok Core, back to 400,000 years before the present day 

- Description of data set: <https://data.ess-dive.lbl.gov/view/doi%3A10.3334%2FCDIAC%2FATG.009>
- data: <https://data.ess-dive.lbl.gov/catalog/d1/mn/v2/object/ess-dive-457358fdc81d3a5-20180726T203952542>

You will use this code to download the data to your computer.

```{r}
download.file("https://data.ess-dive.lbl.gov/catalog/d1/mn/v2/object/ess-dive-457358fdc81d3a5-20180726T203952542",
              destfile = "vostok.icecore.co2")
```

You can click on the file in your Files pane to view it before reading into R.


**Question 10:**

The broad question is: how do recent atmospheric CO2 levels compare to historical levels?

Your answer to Question 10 is going to be a mix of code chunks and text that you put below in "Answer 10:"

- Describe the data set: what are the columns and units? Where do the numbers come from? 

- What is the temporal resolution of the data? 
- Read in and prepare data for analysis.   
- Reverse the ordering to create a chronological record so that each measurement is associcated with calender year.
- Plot data.  
- Combine this time series with the Mauna Loa data (the Warm-up exercise in this assignment).  You will use "bind_rows()".
- Plot the combined data. (the most recent time period must be on the right side of the plot).        
- Describe your conclusions to the question "How do recent atmospheric CO2 levels compare to historical levels?" using your plot as supporting evidence.   

**Answer 10:**

```{r}


```

# Render and committing

Remember to Render your document as HTML and comment+push to GitHub your code and rendered HTML that was created when you knitted the document.  Your GitHub repository should have multiple commits with informative commit messages.

# Attribution

Include citation of any AI-generated assistance or discussion with classmates (per policy in syllabus). Proper documentation of AI-generated assistance includes the prompt, the source (e.g., ChatGPT), and the significant parts of the response.  Proper documentation of discussion with classmates include listing their names and the components discussed.  



